\section{Hints for ``Isomorphisms'' exercises}\label{sec:isomorph:hints} 


\noindent Exercise \ref{exercise:isomorph:iso_4}:  Show a counterexample where the sum of two complex numbers is not the same as the sum of their corresponding ordered pairs.

\noindent Exercise \ref{exercise:isomorph:InvCompIso}(a):  According to Definition~\ref{isomorph_defn}, this involves proving two things about $\phi^{-1}$.  What are they?

\noindent Exercise \ref{exercise:isomorph:InvCompIso}(b): You need to prove the same two things as in part (a).  Use results from the Functions chapter.

\noindent Exercise \ref{exercise:isomorph:GpEquivRel}:   Recall that this involves proving the three properties: reflexive, symmetric, and transitive. You may find that Exercise~\ref{exercise:isomorph:InvCompIso}  is useful.

\noindent Exercise \ref{exercise:isomorph:cyclic_noncyclic}:   Follow Proposition~\ref{not_isomorph_cyclic}.

\noindent Exercise \ref{exercise:isomorph:PhiBij}(a): Use Exercise~\ref{exercise:isomorph:InvCompIso}.~~(b):  Use Proposition~\ref{abelian_non-abelian}. ~~(c):  Use Proposition~\ref{cyclic_noncyclic}. 

\noindent Exercise \ref{exercise:isomorph:proof_thm_7}: For each group property to be proved, use the corresponding group property for $G$ and $H$.


\noindent Exercise \ref{exercise:isomorph:direct_commute}:
 Define a function $\phi:G \times H \rightarrow H \times G$ by:  $\phi(g,h) = \underline{~~~~}$ (you fill in the blank).  Show that this function is in fact an isomorphism. 

\noindent Exercise \ref{exercise:isomorph:exel}(c): Consider 2 cases: (i) 9 divides one of the factors $p_i^{e_i}$ in Proposition~\ref{FactorabelianGroup}; (ii) 9 does not divide any of the factors.

\noindent Exercise \ref{exercise:isomorph:isomex}:  Show that $G \times {\var id}_K$ is a subgroup of $G \times K$, and that $G \times {\var id}_K \cong G$; and similarly for $H$.
\medskip

\textbf{Additional exercises}

\noindent Exercise \ref{exercise:isomorph:eoc}: If you take every other vertex in a hexagon, you get an equilateral triangle. Also note that 180-degree rotation is an element of order 2.

